---
title: "Hitters"
output: rmarkdown::github_document
---
```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

```{r loadPackages, warning=FALSE, message=FALSE, results='hide' }
if(!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, reshape, gplots, ggmap, 
               mlbench, data.table,ISLR, MASS, rpart, rpart.plot, caret,
               randomForest, gbm, tree,leaps)
search()
theme_set(theme_classic())

```


```{r Loading Data}
#Loading Data

raw_hitters.df <- Hitters
head(raw_hitters.df)
str(raw_hitters.df)
```
Q1. Remove the observations with unknown salary information. How many
observations were removed in this process?

```{r ommiting null values}
hitters.df <- raw_hitters.df[!is.na(raw_hitters.df$Salary),]
hitters.df
removed_observations <- length(raw_hitters.df$Salary)-length(hitters.df$Salary)
removed_observations

```
Cleaning the dataset which had a few records seeing missing values in the Salary column saw a total of 59 observations being removed.

Q2.Generate log-transform the salaries. Can you justify this transformation?
```{r log transform}
ggplot(hitters.df, aes(Salary)) + geom_histogram() + ggtitle('Salary without Log tranform')
lgsal =log(hitters.df$Salary)
ggplot(hitters.df, aes(lgsal)) + geom_histogram() + ggtitle('Log tranformed salary')

```
The salary values before the log transformation are right skewed. By log transforming the salary, we are making the salary values somewhat normally distributed.

Q3.Create a scatterplot with Hits on the y-axis and Years on the x-axis using all the
observations. Color code the observations using the log Salary variable. What
patterns do you notice on this chart, if any?
```{r scatter plot}
qplot(hitters.df$Years,hitters.df$Hits, data = hitters.df, main = "Scatterplot for no. of hits with respect to no. of years", colour = lgsal)
```
It is very much clear from the scatter plot that hits are more when the player has experience of around 4 years to 6 years. In the later years, players are not generating as much hits but still are making great salary due to their experience level. Players with 5 or more than 5 are tending to get better salaries as compared to least experienced ones.

Q.4 Run a linear regression model of Log Salary on all the predictors using the entire
dataset. Use regsubsets() function to perform best subset selection from the
regression model. Identify the best model using BIC. Which predictor variables
are included in this (best) model? 

```{r}
set.seed(42)
hitters.lm <- regsubsets(log(hitters.df$Salary) ~ ., data = hitters.df,method = 'exhaustive')
hitters.lm.summary<-summary(hitters.lm)
hitters.lm.summary
hitters.lm.summary$bic


```
The model with the lowest BIC value is said to give the best model. It can be seen from the output as the model with 3 predictor has the lowest BIC value of -159.2777.
The best predictors for predicting the salaries of players are:
Hits, Walks and Years.

Q.5 Now create a training data set consisting of 80 percent of the observations, and a
test data set consisting of the remaining observations. 
 

```{r}
set.seed(42)
train.index <- sample(1:nrow(hitters.df), 0.8 *(nrow(hitters.df))) 
train.df <- hitters.df[train.index, ]
valid.df <- hitters.df[-train.index, ]
```
Q6.Generate a regression tree of log Salary using only Years and Hits variables from the training data set. Which players are likely to receive highest salaries
according to this model? Write down the rule and elaborate on it.

```{r}
set.seed(42)
regtree <- rpart(log(Salary) ~ Years + Hits, data = train.df, method ="anova")
prp(regtree, type = 1, extra = 1, split.font = 2)  
rpart.rules(regtree, cover = TRUE)

```
The player getting highest salary is one who has experience of equal or more than 5 years in the league and has 104 or more hits.
the rule simply is:
Salary = 6.7 when years>=5 and hits>=104.

Q7. Now create a regression tree using all the variables in the training data set.Perform boosting on the training set with 1,000 trees for a range of values of the shrinkage parameter λ. Produce a plot with different shrinkage values on the xaxis and the corresponding training set MSE on the y-axis.

```{r}
set.seed(42)
shrinkage=seq(0.001,0.2,length=100)
terrors <-rep(NA,100)
verrors <- rep(NA,100)
for (i in 1:100) {
  
    boosting <- gbm(log(Salary) ~ ., data = train.df, distribution = "gaussian", 
        n.trees = 1000, shrinkage = shrinkage[i])
    train_pred <- predict(boosting, train.df, n.trees = 1000)
    valid_pred <- predict(boosting, valid.df, n.trees = 1000)
    terrors[i] = mean((log(train.df$Salary) - train_pred)^2)
    verrors[i] = mean((log(valid.df$Salary) - valid_pred)^2)
}
ggplot(data.frame(x=shrinkage, y=terrors), aes(x=x, y=y)) + xlab("Shrinkage Parameter") + ylab("Training MSE") + geom_point()
shrinkage[which.min(terrors)]
```
The shrinkage Parameter having minimum Mean Square error on testing data is 0.2

Q8.Produce a plot with different shrinkage values on the x-axis and the
corresponding test set MSE on the y-axis. 
```{r}
ggplot(data.frame(x=shrinkage, y=verrors), aes(x=x, y=y)) + xlab("Shrinkage Parameter") + ylab("Testing data MSE") + geom_point()
shrinkage[which.min(verrors)]

```

The shrinkage Parameter having minimum Mean Square error on Testing data is 0.08542424

Q9.Which variables appear to be the most important predictors in the boosted
model? 

```{r}
set.seed(42)
boosting <- gbm(log(Salary) ~ ., data = train.df, distribution = "gaussian", 
    n.trees = 1000, shrinkage = shrinkage[which.min(terrors)])
summary(boosting)
```
The important predictors in the boosted model with best shrinkage parameter having minimum MSE are the ones having highest relative influence .
The top 3 important predictors are:
CAtBat
PutOuts
walks

Q 10. Now apply bagging to the training set. What is the test set MSE for this
approach?

```{r}
set.seed(42)
bagging <- randomForest(log(Salary)~., data=train.df, 
                           mtry = 19, importance = TRUE,ntree=1000)
bagging
bagging.prediction <- predict(bagging, valid.df)
mean((bagging.prediction-log(valid.df$Salary))^2)
```
The Test Mean Squared Error for bagging is 0.2487804
